What is an Ancestral Recombination Graph?

### 2022-01-25 paper.tex
Initial theory + Monte Carlo writeup

\linenumbers
\title{What is an Ancestral Recombination Graph?}
% First authors
\author{Author McAuthory}
% Corresponding

@@

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time. This has created exciting
possibilities, with many downstream applications becoming possible based on
these inferred genealogies. The inferred genetic ancestries are usually
referred to as Ancestral Recombination Graphs, or ARGs. Although initially well
defined as a graphical representation of the coalescent with recombination
stochastic process, the interpretation has become unclear as it is now
also understood to represent a particular realisation of a genetic
ancestry.
Inference methods do not all infer the same information:
although all output marginal trees along the genome, there is a great deal of
variation in how much information is inferred and retained in the relationships
between the trees. An important programme of work over the coming years is to
develop and refine ancestry inference methods, assessing the relative strengths
and weaknesses of the various approaches. Using the blanket term ARG hides the
important differences in the approaches, and hampers our ability to compare and
improve inference methods. In this paper we discuss the different approaches we
may use to encode genetic ancestry, and of the differing amounts of information
about the ancestral process that we can represent and hope to observe. We
classify existing inference methods according to the type of ancestry they
infer, and discuss the strengths and weaknesses of the inferred structure for
downstream applications.
\end{abstract}

@@

One major source of confusion is the use of the term "ARG" to refer to both a
backwards-in-time generative process (Griffiths) and a structure which can be
created by that process
\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.

@@

The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}.
The ``big'' ARG~\citep{ethier1990two},
and the ``little ARG'' traversed by
Hudson's algorithm~\citep{hudson1983properties}.

In the theoretical literature, ancestral graphs such as the ARG refer to the
ancestral \emph{process} rather than the outcome of a process.
\begin{itemize}
\item Ancestral Selection
Graph~\citep{krone1997ancestral,neuhauser1999ancestral} models selection
as a graph with virtual lineages [etc etc.]. Importantly, the graph structure
of the ASG does \emph{not} correspond to a realised ancestry, but rather
models possible outcomes.
Other papers possibly worth citing:\citep{etheridge2009coalescent}
\item Ancestral Gene Transfer Graph~\citep{baumdicker2014infinitely}
\end{itemize}

@@

\begin{itemize}
\item Early methods \citep{hein1990reconstructing}
\item Those sampling from the coalescent with recombination
\citep{griffiths1996ancestral,kuhner2000maximum,fearnhead2001estimating},
or SMC
\citep{rasmussen2014genome}
\item Hueristic methods~\citep{song2005efficient,minichiello2006mapping,
parida2008estimating,fallon2013acg,mirzaei2016rent}
\item Discussions of the complexity of inferring ARGs (but we need to
dig into this: is this the complexity of sampling from the coalescent,
finding a ``min ARG'', or something else?
\citep{bordewich2005computational,wang2001perfect,hein1990reconstructing,song2005constructing}
\item Recent methods: Relate, tsinfer, tsdate, etc.
\end{itemize}

>>>>

\linenumbers
\title{What is an Ancestral Recombination Graph?}
% First authors
\author{Author McAuthorface}
% Corresponding

@@

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time. This has created exciting
possibilities, with many downstream applications based on
these inferred genealogies. The inferred genetic ancestries are usually
referred to as Ancestral Recombination Graphs, or ARGs. Although initially well
defined as a graphical representation of the coalescent with recombination
stochastic process, the interpretation has become unclear as it is now
also understood to represent a particular realisation of a genetic
ancestry.
Inference methods do not all infer the same information:
although all output marginal trees along the genome, there is a great deal of
variation in how much information of the relationships between the trees is inferred
and retained. An important programme of work over the coming years is to
develop and refine ancestry inference methods, assessing the relative strengths
and weaknesses of the various approaches. Using the blanket term ARG hides the
important differences in the approaches, and hampers our ability to compare and
improve inference methods. In this paper we discuss the different approaches we
may use to encode genetic ancestry, and of the differing amounts of information
about the ancestral process that we can represent and hope to observe. We
classify existing inference methods according to the type of ancestry they
infer, and discuss the strengths and weaknesses of the inferred structure for
downstream applications.
\end{abstract}

@@

One major source of confusion is the use of the term ``ARG" to refer to both a
backwards-in-time generative process (Griffiths) and a structure which can be
created by that process
\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.

@@

The simplest prominent example of genetic ancestry generated by a stochastic process
is the coalescent process \citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}, which describes the random
ancestral tree at one non-recombining locus obtained by sampling a relatively
small number of individuals at random from a large population.

In a recombining organism, the genetic ancestry of a longer segment of DNA is 
no longer described by a single tree. Instead, ancestral relationships in a
sample can be represented by a sequence of correlated, marginal trees along the
genome. The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing 
stochastic process whose realisations are graphs which into which all of the
marginal ancestral trees are embedded. Coalescences encode common ancestry
as they do in the coalescent, while branching events encode points at which a
marginal tree changes from one position along the genome to the next.

There are several formulations of the ARG, which vary by
the amount of extra information they capture beyond a minimal representation
of all marginal trees. Two prominent examples are the ``big''
ARG~\citep{ethier1990two}, which has a particularly simple specification as a
stochastic process but is computationally infeasible, and the ``little ARG''
traversed by Hudson's algorithm~\citep{hudson1983properties}, which
has a more complex state space and dynamics but can be implemented with 
a more manageable runtime and memory footprint. An efficient simulation
implementation has become available only recently \citep{kelleher2016efficient},
prior to which state-of-the-art simulators were based on the so-called sequentially
Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
performance gains by disallowing mergers which would result in trapped
non-ancestral material. An alternative restricted model is
\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
tree and disallows recombination in any other marginal tree.

Similar graph encodings have also been introduced to model other mechanisms
which change the local ancestral tree along the genome, such as gene
conversion~\citep{wiuf2000coalescent} and
horizontal gene transfer~\citep{baumdicker2014infinitely}.

The idea of using branching-coalescing stochastic processes 
to model genetic ancestral trees has also found application in natural selection.
The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses dynamics identical to the ``big" ARG to simulate an ensemble of correlated
potential ancestral trees. Weak genic selection is incorporated by sampling a 
true ancestry from the ensemble in a non-uniform way.
There is no ASG analogue of the more computationally tractable ``little" ARG,
though some gains in tractability can be made by considering typed 
lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
techniques when recurrent mutation is present\citep{fearnhead2001perfect}.
Extensions of the ASG have been developed to frequency-dependent 
selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
loci whereupon branching is due to both selection and
recombination~\citep{donnelly1999genealogical}, and high fecundity 
reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.

In the theoretical literature, ancestral graphs such as the ARG refer to the
ancestral \emph{process} rather than the outcome of a process.

@@

The ARG has been used as a model for recombining DNA sequence data essentially
since it was first formulated. Early methods focused on finding parsimonious
ARGs consistent with a given set of
sequences \citep{hein1990reconstructing, lyngso2005minimum}.
However, identifying the most parsimonious ARG for a given sequence data set
is NP-hard \citep{wang2001perfect}, and subsequent methods have resorted to
various heuristics \citep{fallon2013acg, hein1993heuristic, ignatieva2021kwarg, 
kelleher2019inferring, minichiello2006mapping, mirzaei2017rent,
parida2008estimating, song2005efficient, thao2019hybrid, zhang2021biobank}. 
The scalability of these methods varies substantially based on the heuristic in use
and the underlying data structure, but the most efficient methods can be practically
used on megabase scale data and tens of thousands of samples under human-like
parameters \citep{kelleher2019inferring, zhang2021biobank}.

An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling 
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference} 
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.
These methods invariably operated on a representation of the ``little ARG", typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.

Due to the computational expense of sampling ``little ARGs", state-of-the-art
Monte Carlo methods typically rely on cheaper, approximate models.
The most prominent examples are \texttt{ARGWeaver} \citep{rasmussen2014genome}
and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
discretisation approximation, but scales to dozens of mammal-like genomes.
The \texttt{ClonalOrigin} method \citep{didelot2010inference,
medinaaguayo2020speeding} uses the \texttt{ClonalFrame}
model and data structure, though it is not as scalable as \texttt{ARGWeaver}.

A very recent MCMC method called \texttt{ARGInfer} has resulted in the first
improvement in the scalability of Monte Carlo methods for the exact ARG model 
by making use of the tree sequence representation of marginal trees, suitably enriched
to facilitate the computations needed for MCMC sampling \citep{mahmoudi2021inference}.
While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends
the feasible range of sequence lengths by at least an order of magnitude when
compared with methods which sample realisations of the exact ``little ARG".
